The effect of linking assumptions and number of response options on inferred scalar implicature rate

### 2018-05-07 writeup.Rmd

On the traditional view of the link between implicature and behavior in sentence verification tasks, scalar implicature is conceptualized as a binary, categorical affair - that is, an implicature is either ‘calculated’ or it isn’t, and the behavioral reflexes of this categorical interpretation process should be straightforwardly observed in experimental paradigms. This assumption has concerning implications for how we must approach analysis of variation in behavior on a truth value judgment task; for example, why did the majority of respondents in the binary condition of our experiment answer “Right” to an utterance of cat or dog when the card had both a cat and a dog on it?

To explain the data on the traditional view, are forced to say that a) not all participants calculated the implicature; or that b) some participants who calculated the implicature did not choose the anticipated response (i.e. “Wrong”) due to some other cognitive reflex which ‘overrode’ the implicature; or some mixture of (b) and (c). We might similarly posit that one or both of these factors underlie the variation in the ternary, quatenary, and quinary conditions (e.g. why were participants roughly split between “Right” and “Kind of right” when the utterance was cat or dog and the card had a cat and a dog?). However, the best we can hope for on this approach is an analysis which traces the general qualitative patterns in the data.

We contrast the above view of implicature and its behavioral reflexes with an alternative linking hypothesis which assumes that participants' behavior can be represented using the model of a soft-optimal pragmatic speaker in the RSA framework. This alternative linking hypothesis contrasts with the traditional view in it is rooted in a quantitative formalization of pragmatic competence which provides us a continuous measure of pragmatic reasoning. Recall that in RSA, pragmatically competent listeners are modeled as a continuous probabilistic distribution of possible meanings given an utterance which that listener hears. The probability with which this listener $L_1$ ascribes a meaning s to an utterance u depends upon a prior probability distribution of potential states of the world $P_w$, and upon reasoning about the communicative behavior of a speaker $S_1$. $S_1$ in turn is modeled as a continuous probabilistic distribution of possible utterances given an intended state of affairs the speaker intends to communicate. This distribution is sensitive to a rationality parameter $\alpha$, the production cost $C$ of potential utterances, and a representation of a literal listener $L_0$ whose interpretation of an utterance is in turn a function of that utterance's truth conditional content $[[u]](s)$ and her prior beliefs about the state of the world $P_w(s)$.

>>>>

On the traditional view of the link between implicature and behavior in sentence verification tasks, scalar implicature is conceptualized as a binary, categorical affair - that is, an implicature is either ‘calculated’ or it isn’t, and the behavioral reflexes of this categorical interpretation process should be straightforwardly observed in experimental paradigms. This assumption raises concerns for analyzing variation in behavior on a truth value judgment task; for example, why did the majority of respondents in the binary condition of our experiment answer “Right” to an utterance of the underinformative "There is a cat or dog" when the card had both a cat and a dog on it?

To explain the data on the traditional view, we are forced to say that a) not all participants calculated the implicature; or that b) some participants who calculated the implicature did not choose the anticipated (i.e., "Wrong") response due to some other cognitive process which overrode the 'correct' implicature behavior; or some mixture of (a) and (b). We might similarly posit that one or both of these factors underlie the variation in the ternary, quaternary, and quinary conditions. However, the best we can hope for on this approach is an analysis which traces the general qualitative patterns in the data. **XXX Brandon, can you be more specific about what you mean by 'tracing a general qualitative pattern'?**

We contrast the above view of implicature and its behavioral reflexes with an alternative linking hypothesis. Recent developments in the field of probabilistic pragmatics have demonstrated that pragmatic production and comprehension can be captured within the Rational Speech Act (RSA) framework, which treats listeners and speakers as agents that reason about each other and choose interpretations and utterances, respectively, that soft-maximize utility [@Frank2012; @Goodman2013; @Goodman2016; @Franke2016; @Qing2014; @Degen2013; @DegenTG2015; @Bergen2016; @Kao2014; @Scontras2017]. We take inspiration from this approach and treat participants' behavior in our experimental tasks as the result of a soft-optimal pragmatic speaker in the RSA framework. This alternative linking hypothesis contrasts with the traditional view in that it is rooted in a quantitative formalization of pragmatic competence which provides us a continuous measure of pragmatic reasoning. In RSA, pragmatically competent listeners are modeled as a continuous probabilistic distribution over possible meanings (states of the world) given an utterance which that listener observes. The probability with which this listener $L_1$ ascribes a meaning $s$ to an utterance $u$ depends upon a prior probability distribution of potential states of the world $P_w$, and upon reasoning about the communicative behavior of a speaker $S_1$. $S_1$ in turn is modeled as a continuous probabilistic distribution over possible utterances given an intended state of the world the speaker intends to communicate. This distribution is sensitive to a rationality parameter $\alpha$, the production cost $C$ of potential utterances, and the informativeness of the utterance, quantified via a representation of a literal listener $L_0$ whose interpretation of an utterance is in turn a function of that utterance's truth conditional content $[[u]](s)$ and her prior beliefs about the state of the world $P_w(s)$.
